---
title: "Experiment 3 - seed germination"
author: "Marcel Dorken"
output: html_document
---

# Analysis of seed-germination success among cattail species in response to shading and leachate (Szabo et al. 2017)

This script relates to the analysis of seed germination success from a greenhouse experiment described in Szabo et al. (2017) in which seeds were placed in water-filled petrie dishes and subjected to a combination of shading and leaf-litter leachate treatments, each with two levels. 

The data is stored as a .csv file and is located in the folder that accompanies this script. This script shows the steps used to evaluate model assumptions and the analyses and figures presented in the paper.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This first block loads all libraries and data used in the analyses.

```{r startup, message=FALSE}
rm( list = ls() )

require( lme4 ) # version 1.1.12
require( nlme ) # version 3.1-128
require( car )

TEx3 <- read.csv("Exp3.Total.csv", header=T) #reads data from *T*ypha *Ex*periment *3* 

```

###Data exploration and model examination 
The first step is data exploration, starting with some dotplots

```{r data exploration 1, echo = FALSE}

TEx3 <- TEx3[order(TEx3$Species),] #sort the data for a nicer dotplot

par(mfrow=c(1,1))

dotchart(TEx3$Seeds.G/TEx3$Seeds.T, #proportion of seeds that germinated
         groups = factor(TEx3$Species), #groups the dots according to species
         xlab="Proportion of seeds", 
         ylab="Species", 
         main="", 
         pch=as.numeric(TEx3$Species))

TEx3 <- TEx3[order(TEx3$Leachate),] #sort the data for a nicer dotplot

dotchart(TEx3$Seeds.G/TEx3$Seeds.T, 
         groups = factor(TEx3$Leachate), 
         xlab="Proportion of seeds", 
         ylab="Leachate", 
         main="", 
         pch=as.numeric(TEx3$Leachate))


```

There are no clear patterns here, either for species or leachate effects, also no evidence for outliers.  

Now for some box plots:

```{r data exploration 2, echo = FALSE}

par(mfrow=c(2,2))

boxplot(Seeds.G/Seeds.T ~ factor(Species), data=TEx3, main = "Species",
        ylab="Proportion of seeds", xlab="Species")

boxplot(Seeds.G/Seeds.T ~ factor(Leachate), data=TEx3, main = "Leachate",
        ylab="Proportion of seeds", xlab="Leachate")

boxplot(Seeds.G/Seeds.T ~ factor(Shade), data=TEx3, main = "Shade",
        ylab="Proportion of seeds", xlab="Shade")

```

* The boxplots suggest that seed germination for *T. latifolia* was lower than for *T. angustifolia* and seed germination for *T. x glauca* was more variable than for the other species - something that was apparent from the dotplots, too. 
* The boxplots further suggest that leachate suppressed seed germination

Run a linear model and test for the importance of the random effect (PlantID)

```{r random effects test}

pS <- TEx3$Seeds.G/TEx3$Seeds.T

Form <- formula(pS ~ Species * Shade * Leachate)
M.gls <- gls(Form, data=TEx3)
M.lme <- lme(Form, random =~1|PlantID, data=TEx3, method = "REML")
anova(M.gls, M.lme) #AIC is a lot lower for the mixed model; random effect is needed 


```

```{r random effects plot, echo = FALSE}

M.lm <- lm(Form, data = TEx3)
par(mfrow=c(2,2))
E <- resid(M.lm) 
Fi <- fitted(M.lm)
plot(Fi, E, xlab="Fitted values", ylab = "Residuals")
boxplot(E ~ Species, data=TEx3, main = "Species") # some heterogeneity of residuals among species levels
boxplot(E ~ Shade, data=TEx3, main = "Shade")  
boxplot(E ~ Leachate, data=TEx3, main = "Leachate") 

par(mfrow=c(1,1), mar=c(5,2,2,1)+0.1) #reset graphical parameters
boxplot(E ~ PlantID, data=TEx3, axes=FALSE, ylim=c(-0.9, 0.4))
abline(0,0); axis(2)
text(1:nlevels(TEx3$PlantID), -0.8, levels(TEx3$PlantID), cex=0.75, srt=90)
```

This figure further supports the need to include the random effect in any analysis of seed germination. The question is whether to use a glmm or lme? The assumptions of the linear model appear to be more-or-less satisfied. Perhaps start with that and then compare with a glmm.

### Model simplification & validation - linear mixed model
```{r model simplification, echo=FALSE}

M1.0 <- lme(Form, random = ~1|PlantID, data = TEx3, method = "ML")
M1.1 <- update(M1.0, .~. -Species:Shade:Leachate)
anova(M1.0, M1.1) # remove the 3-way term

Form2 <- formula(pS ~ Species * Shade + Species * Leachate + Shade * Leachate)
M2.0 <- lme(Form2, random = ~1|PlantID, data = TEx3, method = "ML")
M2.1 <- update(M2.0, .~. -Species:Shade)
M2.2 <- update(M2.0, .~. -Species:Leachate)
M2.3 <- update(M2.0, .~. -Shade:Leachate)
#all interaction terms are significant; run a final model with REML

```

```{r final model, message=FALSE}
M2.final <- lme(pS ~ Species * Shade + 
                  Species * Leachate + 
                  Shade * Leachate, 
                random = ~1|PlantID, data = TEx3, method = "REML")
summary(M2.final)
Anova(M2.final, Type = "III")

```

```{r model assumptions, echo=FALSE}

E.lm <- resid(M2.final, type = "normalized")
F.lm <- fitted(M2.final)
op <- par(mfrow=c(2,2), mar=c(4,4,3,2))
MyYlab <- "Residuals"
plot(F.lm, E.lm, xlab="Fitted values", ylab=MyYlab) # no pattern in the residuals, which is reassuring
boxplot(E.lm ~ Species, data=TEx3, main = "Species", ylab=MyYlab) # these look ok too
boxplot(E.lm ~ Shade, data=TEx3, main = "Shade Treatment", ylab=MyYlab)
boxplot(E.lm ~ Leachate, data=TEx3, main = "Leachate Treatment", ylab=MyYlab)

```

The model assumptions appear to have been met but perhaps a glmm would fit just as well, and better reflect the structure of the data.

```{r glmer test, echo=FALSE}

y <- cbind(TEx3$Seeds.G, TEx3$Seeds.T-TEx3$Seeds.G)

M1.g.0 <- glmer(y ~ Species * Shade * Leachate + (1|PlantID), family = "binomial", data = TEx3)
M1.g.1 <- update(M1.g.0, .~. -Species:Shade:Leachate)
anova(M1.g.0, M1.g.1) 

summary(M1.g.0)
Anova(M1.g.0)

```

Ugh! the 3-way interaction is significant. Model validation is more difficult to assess with the glmm; the safest approach is probably to use the linear model. At least it is interpretable.

###Final model plot
```{r plot linear model, message=FALSE}
par(op)
par(mar=c(4,4,2,2))

boxplot(pS ~ Species + Shade, data = TEx3, at=c(1,1.8,2.6, 4.5,5.3,6.1), 
        xaxt='n', col=c("greenyellow", "green", "green4"), boxwex=0.6,
        ylab = "Proportion germinated")
mtext(side = 1, at = c(1.8, 5.3), text = c("Unshaded", "Shaded"), line = 2.5)
mtext(side = 1, at = c(1,1.8,2.6,4.5,5.3,6.1), text = c("TA", "TG", "TL"), line = 1)

boxplot(pS ~ Species + Leachate, data=TEx3, at=c(1,1.8,2.6, 4.5,5.3,6.1),
        xaxt='n', col=c("greenyellow", "green", "green4"), boxwex=0.6, 
        ylab = "Proportion germinated") 
mtext(side = 1, at = c(1.8, 5.3), text = c("No Leachate", "Leachate added"), line = 2.5)
mtext(side = 1, at = c(1,1.8,2.6,4.5,5.3,6.1), text = c("TA", "TG", "TL"), line = 1)

boxplot(pS ~ Shade + Leachate, data=TEx3, at=c(1,1.8, 3.5,4.3), 
        xaxt='n', col=c("white", "grey66"), boxwex=0.6,
        ylab = "Proportion germinated") 
mtext(side = 1, at = c(1.4, 3.9), text = c("No Leachate", "Leachate added"), line = 2.5)
mtext(side = 1, at = c(1,1.8,3.5,4.3), text = c("Unshaded", "Shaded"), line = 1)

```

###Interpretation

Together with model results, the plot of the species x shade interaction indicates that shade suppressed or had little effect on seed germination of *T. angustifolia* and *T. x glauca*, respectively, but was associated with increased seed germination by *T.latifolia*.

Model results and the plot of the species x leachate interaction indicates that seed germination by *T. angustifolia* and *T. latifolia* were negatively affected by leaf-litter leachate from *T. x glauca* but that *T. x glauca* was not.    

Finally, the shade x leachate effect indicates that the negative effect of leachate was reduced/ameliorated by shading.


